---
title: "Oct 1 Activities: M&Ms; Survey; Sampling"
date: 9-30-2025
Casey Glickman
---

All three class activities are on this .rmd notebook

Activity #1: Activity: M&Ms - Central Limit Theorem

Working in groups, each group will get a handful of candy from the bag, and then count the number of red and black candies. Log those in this Google Sheet 

https://docs.google.com/spreadsheets/d/1KOeoVTq2W7L2T9sWjwh5U_2f59GsTeAylVGD53rNWww/edit?usp=sharing

(give your group a nice name).

Take a look at the values that we all get - what do you notice about the patterns? This is an exercise about the central limit theorem. 

The central limit theorem states that the sampling distribution of the mean of any independent, random variable will be normal or nearly normal, if the sample size is large enough.

Then discuss and try to answer these questions:

--How close is your group's mean to the true proportion of 50%?
--Why do different groups get different results?
--What happens to individual extreme results when we combine all groups?
--How does this relate to polls and surveys in real life?

Write your answer below. And then you may eat the M&Ms.

**A:
Our groups mean is 85% which is much larger than the true proportion of 50%. We had a small number of brown m&m's in our sample so there is a high percentage of colored m&m's

different groups have different results because it is a random sample of m&m's. Some groups had larger samples, so even if two groups had the same number of red m&m's the percentage of the total would be different.

when we combine all the groups the extreme values become less relevant because all of the groups averages get mixed together.

In real life random sampling will not give us an accurate representation of the whole picture. If we were to do random sampling at a polling site you might only collect data from people of a similar political background and therefore your data would have inaccurate results.


**

------------------------------------------------------------------------------

# Activity #2:

Belonging Survey

Working in pairs, take a look back at the main findings from UMD's Belonging Survey:

https://president.umd.edu/internal/belonging


Specifically, focus on the "Sample Description" section and the Appendix. Discuss the differences and between the population and the sample, and what the document says about how the survey attempted to deal with them. Are there any decisions described in the document that raise questions in your minds? Detail the differences you see and your understanding of the methods they employed (it's ok if you don't understand all of them), and add your questions. You should discuss this as a pair but submit individually.

**A:
In the sample, staff members were underrepresented and students were underrepresented compared to the schools actual population. The sample had a larger female representation and a smaller male representation when compared to the schools actual gender breakdown. In adittion, Asian americans were over represented in the sample, compared to black and hispanic students who were under represented. One decision that seems questionable in the data analysis was the decision to group smaller minorities into one category called "Races/Ethnicities with Small Sample Sizes."

**

------------------------------------------------------------------------------

# TASK 3: SAMPLING

Based on the class sampling demo (available on Elms), sample data from a population using R and the Tidyverse. You should get started by loading the tidyverse in a new R block.

### Task 1: Load the tidyverse

**A**
```{r}
library(tidyverse)

```


### Task 2: Load 1,000 rows of data with the ages of people from 18-80

**A**
```{r}
set.seed(42)
younguns <- tibble (
  age = sample(18:80, 1000, replace = TRUE)
)

head(younguns)
```

### Task 3: Sample 100 rows of data from the dataset
**A**
```{r}

sample_100 <- younguns |>
  sample_n(100)

```


### Task 4: Calculate the mean age and standard deviation of the sample
**A**
```{r}

sample_100 |> summarize(mean = mean(age), median = median(age), sd = sd(age))
```


### Task 5: Calculate the mean age and standard deviation of the population
**A**
```{r}
younguns |> summarize(mean = mean(age), median = median(age), sd = sd(age))

```


### Task 6: Compare the sample and population statistics
Compare the means and standard deviations. Are they similar? Go back and re-run the sampling and statistics calculations with a smaller sample size. How do the results change?

**A
The sample has a slightly smaller median and mean, but overall the sample is an accurate representation of the entire population because the sample mean is very close to the population mean. If you were to change the sample size to 50, the mean and standard deviation stay close to the same.
**

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
