---
title: "HW3 WH Salaries"
---
Casey Glickman
In this assignment, we will be working with a dataset containing information about the salaries of White House staffers as of July 2024.

### Task 1: Load the Tidyverse

You will need to load the tidyverse library

```{r}
library(tidyverse)
```

Then you'll load the data from the web. The URL for the data is: https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv. You will create a dataframe called `wh_salaries` that contains the data from this URL.

### Task 2: Load the Data and Describe It

```{r}
wh_salaries <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/biden_wh_salaries_2024.csv") 
```
### Task 3: Describe the Data

You should examine that dataframe carefully by clicking on it in the Environment tab in RStudio. Write a couple of sentences describing the categories of incidents. From a reporting standpoint, what stands out to you about this data? Are there any salary values that will cause problems for analysis?

The salaries are categorized by name, status as either a employee or detailee, salary and job title. The salaries of $0 immediately stood out to me as values that will skew the data and it is confusing that an employee can be making no money. This is defintely a statistic that will cause some problem for analysis. The standardized paymnets also jumped out at me. Most staff are clustered in just a few salary bands ($139,500, $155,000, $168,000, $180,000) which shows how the White House structures their pay.


### Task 4: Calculate the Mean and Median of the Salaries

Write code to calculate the mean and median of the salaries in the dataset using the `summarize()` function. You don't need to make a new dataframe, just display the results.

```{r}
wh_salaries |> 
    summarize(mean_salary = mean(salary),
            median_salary = median(salary),
            mean_salary = mean(salary)
)
```

### Task 5: Add in the minimum and maximum salaries

Take the code you just wrote and add in the minimum and maximum salaries (the functions for that are `min()` and `max()`). Your results should show four columns: mean_salary, median_salary, min_salary, max_salary.

```{r}
wh_salaries |> 
    summarize(mean_salary = mean(salary),
            median_salary = median(salary),
            max_salary = max(salary),
            min_salary = min(salary)
)
```

### Task 6: Describe the Results

Write a couple of sentences describing the results of your analysis. What do the mean and median salaries tell you about the data? What do the minimum and maximum salaries tell you? Are there any outliers in the data?

There are a few high values like $251,258, relative to many clustered mid-range salaries. However there are also a few low outliers like the $0 salaries. The mean salary is slightly lower than the median, which tells us the distribution is skewed downward by the lower outliers in the data set. 

### Task 7: Story Idea

Based on the data, what is your best story idea? What is the next step you would take to develop that story idea?

A  story idea would be comparing employee salaries versus detailees salaries. Detailees from other agencies make up a notable share of the roster and often have different salary levels than permanent White House employees. For the article, it would be interesting to compare average pay and roles of detailees vs. employees, and explore what this says about how much the administration relies on outside agency staff.
