--
title: "HW7_MandMs"
date: 10/1/2025
Casey Glickman
---


In this assignment, you'll be picking up where you left off with the M&Ms activity. You'll load the data and perform some basic statistical analysis. Follow the steps below to complete the assignment.

### Task 1: Load the tidyverse

```{r}
library(tidyverse)
```


### Task 2: Load an M&Ms dataset from https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv
```{r}
mandms <- read_csv("https://raw.githubusercontent.com/dwillis/jour405/refs/heads/main/data/mandms.csv")
```


### Task 3: Sample 5 rows of data from the dataset, and then sample 10 rows, saving each to separate dataframes called sample5 and sample10. Be sure to use set.seed(1234) before you start.
```{r}
set.seed(1234)
sample5 <- mandms |>
  sample_n(5)
sample10 <- mandms |>
  sample_n(10)
```


### Task 4: Calculate the mean, median and standard deviation of red M&Ms from each of the samples.
```{r}
sample5 |> summarize(mean = mean(red), median = median(red), sd = sd(red))
sample10 |> summarize(mean = mean(red), median = median(red), sd = sd(red))
```


### Task 5: Calculate the mean, median and standard deviation of red M&Ms from the population

```{r}
mandms |> summarize(mean = mean(red), median = median(red), sd = sd(red))
```

### Task 6: Make three histograms: one each for sample5, sample10 and the original dataset. Include a headline. Use a consistent binwidth.

```{r}
mandms |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 10) + labs(title = "Total M&Ms",  x = "red M&Ms" ,  y = "count")



sample5 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 10)  + labs(title = "Sample of 5 M&M groups" , x = "red M&Ms" ,  y = "count")

sample10 |> 
  ggplot() + 
  geom_histogram(aes(x = red), binwidth = 10)  + labs(title = "Sample of 10 M&M groups", x = "red M&Ms" ,  y = "count")

```


### Task 7: Answer questions


Compare the three sets of means, standard deviations and histograms. Which are more helpful to you in understanding the accuracy of the samples?
All three sets of data have have similar means which tells us that the sample data was a fairly accurate representation of the entire population. The population data for all of the M&M's also had the highest standard deviation which makes sense since there are more M&Ms the range of data will be larger. 

What does this exercise tell you about the relationship between sample size and estimate accuracy?
This exercise tells us that larger sample sizes are a more accurate representation of the entire population data. This is evident when comparing the population histogram with the sample size 10 histogram. However, this exercise also shows that sample sizes are random
and because of that, the correlation from larger sample sizes to population data is not always accurate. For instance, the sample of 5 M&M groups was closer to the actual mean that the sample of 10 M&M groups.

How does random sampling introduce variation into statistical estimates?
Random sampling cane cause variation in statistical estimates because each sample is only one possible subset of the population. In this example, different random samples will have a different amount of M&Ms and different colors; therefore measures like mean and standard deviation will change depending on the sample.

If you repeated this process, would the samples always be identical? Why or why not?
The samples would not be identical because they are randomly selected. The 5 M&M groups could look completely different, and same goes foe the 10 M&M groups. 

When finished, save your work, switch to GitHub Desktop, then add, commit and push your changes to GitHub and submit the URL of the notebook in ELMS.
