---
title: "Regression and Beers"
author: "Wells"
date: "2025-12-02"
output: html_document
---

This is another exercise to break down how a simple linear regression works and how to use it in real life.
The topic: beer.

This following exercise comes from Diez, D., Cetinkaya-Rundel, M., & Barr, C. (2024). OpenIntro Statistics (4th ed.). http://www.openintro.org/redirect.php?go=os&referrer=os4_pdf, p 335


Many people believe that gender, weight, drinking habits, and many other factors are much more important in predicting blood alcohol content (BAC) than simply considering the number of drinks a person consumed. Here we examine data from sixteen student volunteers at Ohio
State University who each drank a randomly assigned number of cans of beer. These students were evenly divided between men and women, and they differed in weight and drinking habits. Thirty minutes later, a police officer measured their blood alcohol content (BAC) in grams of alcohol per deciliter of blood.



Load libraries
```{r}
library(tidyverse)
```


Load data
```{r}
bac_data <- read_csv("https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/activities/bac%20regression/bac_data.csv")

bac_data

#Student #1 drank 5 beers and has a blood alcohol level of 0.13, etc
```


# Linear Regression
```{r}
# Run the regression
model <- lm(bac ~ beers, data = bac_data)

# Display regression results
cat("\n\nRegression Results:\n")
print(summary(model))
```

 Plot the data and the model line
```{r}
# Create regression equation text
eq_text <- sprintf("BAC = %.4f + %.4f * beers", 
                   coef(model)[1], coef(model)[2])


ggplot(bac_data, aes(x = beers, y = bac)) +
  geom_point(color = "darkblue", size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 1) +
  labs(
    x = "Number of Beers",
    y = "Blood Alcohol Content (BAC)",
    title = "BAC vs Number of Beers Consumed"
  ) +
  annotate("text", x = min(bac_data$beers), y = max(bac_data$bac), 
           label = eq_text, hjust = 0, vjust = 1, size = 4) +
  theme_minimal()

```

**Task 1: Interpret the results of this regression. Answer the following questions:**



**Answer:
a) What the predictor or independent variable? What is the response or dependent variable? 
The independent variable is the number of beers and the dependent variable is the blood alcohol level

b) What is the regression model asking and what does it tell us?
The regression model is asking if there is a correlation between the number of beers consumed and your blood alcohol content

c) Do the data provide strong evidence that drinking more cans of beer is associated with an increase in blood alcohol? 
The data provides evidence of a weak positive correlation. The Adjusted R-squared value is 0.5591 which shows that there is a positive correlation but not a strong one. A strong positive correlation would have an r-squared value of 1.

d) State the null and alternative hypotheses, report the p-value, and state your conclusion.
Null hypothesis: there is no statically significant correlation between the number of beers drank and blood alcohol levels
Alternative hypothesis: As the number of beers consumed increased, blood alcohol levels also increase
The p value is 0.0005239 which is extremely low meaning that the data is highly significant. We can reject the null hypothesis and accept the alternate hypothesis that there is a positive correlation between the number of beers consumed and BAC levels.**



