---
title: "final project 1"
output: html_document
date: "2025-11-12"
Casey Glickman
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```

```{r}
# Read the data
tuition_data <- read_csv("https://raw.githubusercontent.com/wellsdata/Jour405_student_repo_fall_2025/refs/heads/main/project/college_scorecard_2021_2024.csv")
```

###: Analysis questions
What does it contain and what time period does it cover?
This data contains tuition rates from colleges in Maryland and Virginia. It includes in state and out of state tuition rates as well as undergraduate student enrollment. The data starts at the 2021-2022 academic year and covers all the way through the 2023-24 year

Which columns will be most important for your analysis?
For my analysis I will be focusing on recent in state tuition rates to see, on average how these fees vary depending on that state. The "2023-24 tuition fee in" and the "state" columns will be the most important for my analysis so I can compare the average tuition for in state students in Maryland and Virginia.  

What's missing that you might need or want?
This data does not include changes in inflation. It would be interesting to see if tuition rates are increasing to keep up with inflation or if rates are falling behind inflation or increasing at a faster rate. For this analysis I would also want to know if the institution list includes every college in the state of Maryland and Virginia and how these schools break down in terms of public universities, private colleges and community colleges. 

```{r}
tuition_data |>
  summarize(
    #average 2023-2024 tuition for in state students
    mean = mean(`2023_24_tuitionfee_in`, na.rm = TRUE),
    #median 2023-2024 tuition for in state students
    median = median(`2023_24_tuitionfee_in`, na.rm = TRUE),
    #standard deviation for in state 2023-2024 tuition 
    sd = sd(`2023_24_tuitionfee_in`, na.rm = TRUE),
    #lowest tuition for in state 
    min = min(`2023_24_tuitionfee_in`, na.rm = TRUE),
    #highest tuition for in state
    max = max(`2023_24_tuitionfee_in`, na.rm = TRUE)
  )
```
### Analysis questions 
Next, propose a newsworthy hypothesis (or several of them) for your dataset. You'll be looking to find (or eliminate possible story ideas) using the techniques you've learned in this class. You'll start that effort this week by coming up with some theories or questions that you're interested in exploring that can be answered with the data set you selected

Newsworthy hypotheses
1. Maryland community colleges have cheaper in state tuition rates compared to Virginia's community college in state tuition
2. Tuition rates have increased in Maryland and Virginia over the past three years
3. As tuition rates increase across colleges, enrollment rates decrease 
4. Colleges with smaller enrollments have higher tuition rates

Null hypothesis: There is no statically significant difference in average 2023-24 in state tuition for colleges in Maryland versus colleges in Virginia 
Alternate hypothesis: There is a statistically significant difference in average 2023–24 in-state tuition between colleges in Maryland and Virginia.

What questions do you have about it or need help answering?
I need help figuring out how to navigate colleges that don't have tuition rates listed. Many colleges with low enrollment have N/A listed for their tuition rates, so I will need to figure out how to work around the gaps in this data 


```{r period-means}
# Categorize data to show average 2023-24 in state tuition in Maryland and Virginia  
period_rates <- tuition_data |>
  group_by(state) |>
  summarize(
    mean_rate = mean(`2023_24_tuitionfee_out`, na.rm = TRUE),
    se = sd(`2023_24_tuitionfee_out`, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  ) |>
  mutate(
    lower_ci = mean_rate - qt(0.975, n() - 1) * se,
    upper_ci = mean_rate + qt(0.975, n() - 1) * se
  )

# Create a bar chart to visualize the difference in average tuition rates by state
ggplot(period_rates, aes(x = state, y = mean_rate, fill = state)) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(ymin = lower_ci, ymax = upper_ci), width = 0.2) + scale_y_continuous(limits = c(0, 40000)) +
  labs(title = "Maryland and Virginia 2023-2024 In State College Tuition",
       x = "State", y = "Mean 2023-2024 in state tuition ") 
```


```{r anova-test}
# Run one-way ANOVA
anova_result <- aov(`2023_24_tuitionfee_in` ~ state, data = tuition_data)
summary(anova_result)
```
```{r}

# Perform t-test comparing Maryland and Virgina
t_test_periods <- t.test(  `2023_24_tuitionfee_in` ~ state, data = tuition_data)

# Display results
t_test_periods
```

### Interperting the Results 
In the ANOVA test we found that the p-value was .945. Since this value is larger than .05 we know that the difference in the average tuition rates is likely due to random chance and we therefore fail to reject the null hypothesis. In the t-test the p-value was .951 which again is very large and means we fail to reject the null hypothesis. The test statistic is used to determine if the difference between the mean of two groups is statistically significant or due to random chance. A larger t-statistic means the difference is more likely to be real and not just due to random chance. However, the two sample t-test found that the t-statistic was .061 which is very small and indicates that the difference between the means is insignificant. Overall, these statistical tests can tell us that there is no statically significant difference in the average 2023-24 in state tuition rates between Maryland and Virginia Therefore we have to accept the null hypothesis and reject the alternate hypothesis. 